---
title: "Project 1"
author: "Isaias Sanchez Saavedra, Jared Woolridge, Meadow Olivas"
date: "2026-04-01"
output: html_document
---


# Contents

## Background .......................................................... 1

## Data ................................................................ 1

## Project Objectives .................................................. 2

## Objective 1 ......................................................... 2

## Objective 2 ......................................................... 2

## Objective 3 ......................................................... 2

## Objective 4 ......................................................... 2

## GitHub Log .......................................................... 2


# Background

The World Health Organization has recently employed a new data science initiative, CSIT-165, that uses data science to characterize pandemic diseases. CSIT-165 disseminates data driven analyses to global decision makers.

CSIT-165 is a conglomerate comprised of two fabricated entities: Global Health Union (GHU) and Private Diagnostic Laboratories (PDL). Your and your partner’s role is to play a data scientist from one of these two entities. 

# Data 

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE). Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. Country/region are countries/regions hat conform to World Health Organization (WHO). Lat and Long refer to coordinates references for the user. Date fields are stored in MM/DD/YYYY format.

```{r,echo=FALSE}
# Download live data
conf_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"
death_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"

confirmed <- read.csv(conf_url)
deaths <- read.csv(death_url)
```

# Project Objectives 

## Objective 1
```{r,echo=TRUE}
# get first day values (column 5 is first date)
first_day_confirmed <- confirmed[,5]
first_day_deaths <- deaths[,5]

# find index of max values
max_confirmed_index <- which.max(first_day_confirmed)
max_deaths_index <- which.max(first_day_deaths)

# get location of those max values
origin_confirmed <- confirmed[max_confirmed_index, c("Province.State","Country.Region")]
origin_deaths <- deaths[max_deaths_index, c("Province.State","Country.Region")]

# check if both match using if statement
if(origin_confirmed$Country.Region == origin_deaths$Country.Region){
  
  # save origin
  origin <- origin_confirmed
  
  # print result
  print(paste("Origin is:", origin$Province.State, origin$Country.Region))
}
```


## Objective 2
```{r,echo=TRUE}

# get last column (latest date)
latest_col <- ncol(confirmed)

# store the most recent first case day found so far
most_recent_day <- 0

# list to store regions
recent_regions <- c()

# loop through each region
for(i in 1:nrow(confirmed)){
  
  # get all date values for that region
  row_data <- confirmed[i, 5:latest_col]
  
  # find days with cases
  case_days <- which(row_data > 0)
  
  # only continue if region has at least one case
  if(length(case_days) > 0){
    
    # get first day with a case
    first_case_day <- case_days[1]
    
    # check if this is the most recent so far
    if(first_case_day > most_recent_day){
      
      # update most recent day
      most_recent_day <- first_case_day
      
      # reset list and store this region
      recent_regions <- c(confirmed$Country.Region[i])
      
    } else if(first_case_day == most_recent_day){
      
      # if tie, add to list
      recent_regions <- c(recent_regions, confirmed$Country.Region[i])
    }
  }
}

# print result
print(recent_regions)
```
## Objective 3


```{r,echo=FALSE}

# Load the library 
library(geosphere)

```

```{r,echo=TRUE}
# get coordinates of origin
origin_coords <- c(confirmed$Long[max_confirmed_index], confirmed$Lat[max_confirmed_index])

# loop through recent regions
for(region in recent_regions){
  
  # get that region data
  subset_data <- subset(confirmed, Country.Region == region)
  
  # get coordinates
  coords <- c(subset_data$Long[1], subset_data$Lat[1])
  
  # calculate distance in meters
  distance_meters <- distm(origin_coords, coords)
  
  # convert meters to miles
  distance_miles <- distance_meters * 0.000621371
  
  # print result
  print(paste(region, "is", round(distance_miles,2),
              "miles away from", origin$Province.State, origin$Country.Region))
}
```
The distance between the most recent region and the origin of the first confirmed COVID-19 cases was calculated using geographic coordinates. The R programming language function distm() from the geosphere package was used to measure the distance in meters between the two locations. This value was then converted into miles to make the result easier to understand. The output shows that the United Kingdom is 9033.32 miles away from Hubei, China, which was identified as the origin. This result demonstrates how far the virus spread geographically from its initial outbreak location.

##Objective 4 
```{r, echo=TRUE}

# get latest values
confirmed_latest <- confirmed[, ncol(confirmed)]
deaths_latest <- deaths[, ncol(deaths)]

# remove cruise ships (Lat and Long not equal to 0)
valid_rows <- confirmed$Lat != 0 & confirmed$Long != 0

# create cleaned data
data <- data.frame(
  Country = confirmed$Country.Region[valid_rows],
  Confirmed = confirmed_latest[valid_rows],
  Deaths = deaths_latest[valid_rows]
)

# calculate risk score
data$Risk <- 100 * (data$Deaths / data$Confirmed)

# remove NA values
data <- data[!is.na(data$Risk), ]

# find highest risk
highest_risk <- data[which.max(data$Risk), ]

# find lowest risk (if tie → most confirmed)
min_risk <- min(data$Risk)
lowest_candidates <- data[data$Risk == min_risk, ]
lowest_risk <- lowest_candidates[which.max(lowest_candidates$Confirmed), ]

# global risk score
global_risk <- 100 * sum(data$Deaths) / sum(data$Confirmed)

# print results
print(highest_risk)
print(lowest_risk)
print(paste("Global risk score:", round(global_risk,2)))
```
The area with the lowest risk score is North Korea, meaning relatively few deaths occurred compared to confirmed cases. The area with the highest risk score is China, where the higher proportion of confirmed cases resulted in death. Compared to the global risk score, these regions show that COVID-19 severity varies widely between locations. Risk scores are useful because they quickly highlight areas that may need more medical resources, but they have limitations. Risk scores assume all cases are detected and do not account for population differences, healthcare quality, or reporting accuracy. Therefore, while helpful, risk scores should be interpreted alongside other information to fully understand the risk level.


## Objective 5
```{r, echo=TRUE}
#packages needed to make a table and use dplyr
library(dplyr)
library(knitr)

#getting the values of deaths and cases
confirmed_cases <- confirmed[,ncol(confirmed)]
confirmed_deaths <- deaths[,ncol(deaths)]

#linking the cases with the country
country_confirmed <- data.frame(
  Country = confirmed$Country.Region[valid_rows],
  coun_con = confirmed_cases[valid_rows]
  )

#linking the deaths with the country
 country_deaths <- data.frame(
   countryd = deaths$Country.Region[valid_rows],
   coun_dea = confirmed_deaths[valid_rows]
   )
 
 #using dplyr to group the different regions of a country and sorting the confirmed cases by country and taking only the top 5 highest values
 total_confirmed <- country_confirmed %>%
   group_by(Country) %>%
   summarize(amnt_con=sum(coun_con, na.rm = TRUE))%>%
   arrange(desc(amnt_con))%>%
   slice(1:5)
 
 ##using dplyr to group different regions of a county and sorting the deaths by country and taking the 5 countries with the hihgest values
 total_deaths <- country_deaths %>%
   group_by(countryd) %>%
   summarize(amnt_deaths=sum(coun_dea, na.rm =TRUE))%>%
   arrange(desc(amnt_deaths))%>%
   slice(1:5)
 
 # 
 # country_total <- data.frame(
 #   Country = total_confirmed[1],
 #   Cases = total_confirmed[2],
 #   Deaths = total_deaths[2]
 # )
 # 
 
 # The tables for the top 5 countries with confirmed cases and the top 5 countries with confirmed deaths.
 kable(total_confirmed, col.names = c("Country", "Confirmed Cases"))
 kable(total_deaths, col.names = c("Country", "Confirmed Deaths"))
 
 
```


